---
title: "Seasonal Package Primer with Hands-on Examples"
author: "Brian C. Monsell"
date: "4/7/2022"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Outline

- Seasonal Package Basics
- Running One Series
- Running Multiple Series
- Examples
- Resources

## I expect that you 

- Know a little about R 
  - lists and time series (`ts`) objects
- Know a little about running X-13ARIMA-SEATS
  - input and output files
  
## Seasonal Package

- Run X-13ARIMA-SEATS within R
- Data input
  - Time series objects
  - Lists or matrices of time series objects
- Output 
  - incorporated into the output of the package

## X-13ARIMA-SEATS Input Files

- Data Files
- Input Specification Files
- User defined Regression and Prior Adjustment Factors
- Metafiles
  
## X-13ARIMA-SEATS Output Files

- Output
- Graphics
- Saved tables
- Diagnostics (UDG and Log)
- Error

## X-13ARIMA Spec File

```
series{
  title="International Airline Passengers "
  start=1949.01   file = "airline.txt"
}
transform{ function=auto }
estimate{ }
regression{ aictest=(td easter) }
arima{  model = (0 1 1)(0 1 1)  }
outlier{ }
x11{ save = (d10 d11 d12 d13 d16 e18) }
spectrum{ }
```

## Example

```{r air_example, echo = TRUE}
# run Airline Series, 
# X-11 seasonal adjustment

m_air <- 
    seasonal::seas(AirPassengers, x11="", 
                   arima.model="(0 1 1)(0 1 1)")
```
## Equivalent spec file

```
 series{
   title = "iofile"
   file = "C:\Users\MONSEL~1\AppData\Local\Temp\RtmpE3tzrP\x1332243080443e/iofile.dta"
   format = "datevalue"
   period = 12
 }
  
 transform{
   function = auto
   print = aictransform
 }
  
 regression{
   aictest = (td easter)
 }
  
 outlier{
  
 }
  
 x11{
   save = (d10 d11 d12 d13 d16 e18)
 }
  
 arima{
   model = (0 1 1)(0 1 1)
 }
  
 estimate{
   save = (model estimates residuals)
 }
  
 spectrum{
   print = qs
 }
```

## Specifying Arguments in 'seas'

To set a specific argument from an X-13ARIMA-SEATS specification, use 

```
spec.argument = value
```

Examples:

```
arima.model = “(0 1 1)(0 1 1)”
forecast.maxlead = 60
x11.siglim = c(1.5, 3.0)
```

## Default settings for `seas`

- Automatic transformation identification
- Automatic model identification
- Automatic outlier identification
- AICC tests for trading day, Easter regressors
- SEATS seasonal adjustment

## Override default settings

- Specify transformation 
  - `transform.function = “log”`
- Set `outlier = NULL`
- Set `regression.aictest = NULL`
- Specify an ARIMA model
  - `arima.model = “(0 1 1)(0 1 1)”`
- Specify an X-11 argument or set `x11 = “ “`

## Difference between NULL and " "

`outlier = “ “` \
Use the empty spec – `outlier{ }`

`outlier = NULL` \
Do not include `outlier` spec in run

## Read in data from an external file

- `import.ts` function
  - Reads data from external files in datevalue format (default) 
  - Output is a ts object

## import.ts example

```{r UN_example, echo = TRUE}
# read unemployed males 
n3000013 <-
   seasonal::import.ts("data/n3000013.dat")

n3000013
```

## Seasonally adjusting one series

- `seas` function
  - Generates the seasonal adjustment based on the options provided
- Output is a seas object, a list object that contains 
  - input
  - diagnostics
  - output
  
## seas example

```{r UN_seas, echo = TRUE}
# seasonally adjust n3000013
n3000013_seas <- seasonal::seas(n3000013, 
    series.span = "2003.1,",
    arima.model = "(0 1 1)(0 1 1)", forecast.maxlead = 60, 
    estimate.exact = "arma", estimate.maxiter = 10000, 
    check.print = c("none", "+acf", "+acfplot", 
                    "+normalitytest"), 
    regression.aictest = NULL, outlier.types = "all",
    regression.variables = c("ls2008.may", "tc2020.apr"))
```

## out function

- Shows output from seasonal adjustment run in a browser

```{r UN_out, echo = TRUE}
seasonal::out(n3000013_seas)
```

## Structure of the seas object

```{r UN_seas_structure, echo = TRUE}
names(n3000013_seas)
```

## Seas Functions: Data Extraction

- `seasonal::final`
  - Extracts final seasonal adjustment

- `seasonal::trend`
  - Extracts final trend component

- `seasonal::series`
  - Extracts saved series from seas object
  - Specify either a table name (`“seasonal”`) or table abbreviation (`“d10”`)  
  - If table not in the seas object, 'seas' will be reevaluated
- List of all importable tables from X-13ARIMA-SEATS
  - `?seasonal::series`
  
## Seas Functions: Data Extraction

```{r UN_extract, echo = TRUE}
# extract final seasonal adjustment
n3000013_sa    <- seasonal::final(n3000013_seas)
n3000013_trend <- seasonal::trend(n3000013_seas)
head(cbind(n3000013_seas$x, n3000013_sa, n3000013_trend))

# extract SEATS seasonal factors
n3000013_sf    <- seasonal::series(n3000013_seas, "s10")
head(n3000013_sf)
```

## Plot of Seasonal Factors

```{r UN_monthplot, echo = TRUE, dev='png'}
monthplot(n3000013_sf)
```

## Plot of ori, SA, trend

```{r UN_plot, echo = TRUE, dev='png'}
this_x     <- window(n3000013_seas$x, start = c(2016, 1))
this_sa    <- window(n3000013_sa, start = c(2016, 1))
this_trend <- window(n3000013_trend, start = c(2016, 1))
plot(this_x, main = "n3000013: SA and Trend", ylab = "n3000013",
     col = "grey")
lines(this_sa, col = "blue")
lines(this_trend, col = "brown")
```

## Seas Object: Diagnostics

- `n3000013_seas$udg`
  - Contains output from X-13ARIMA-SEATS `udg` file
  - “Unified Diagnostics File”
  - `keyword: value` structure

## Example UDG File

```
date: Sep 22, 2016  
time: 09.07.36 
version: 1.1
build: 34
output: html
srstit: X-13ARIMA-SEATS run of airline
srsnam: airline                                
freq:    12
span:  1st month,1949 to 12th month,1960
constant:     0.0000000000E+00
transform: Log(y)
nfcst:    60
ciprob:     0.950000
lognormal: no
mvval:     0.1000000000E+10
iqtype: ljungbox
samode: multiplicative seasonal adjustment
```

## udg Function 

```{r UN_udg1, echo = TRUE}
names(seasonal::udg(n3000013_seas))
```
## udg Function: Examples

```{r UN_udg2, echo = TRUE}
seasonal::udg(n3000013_seas,"nfcst")

seasonal::udg(n3000013_seas,"spectype")

seasonal::udg(n3000013_seas,"qsori")
as.vector(seasonal::udg(n3000013_seas,"qsori"))

n3000013_seas$udg[["nreg"]]
seasonal::udg(n3000013_seas,"nreg")
```


## Import spec files: import.spc

```{r UN_import1, echo = TRUE}
# import seasonal options from specification file

n3000013_import_spc <-  
    seasonal::import.spc("n3000013.spc")

names(n3000013_import_spc)
```

## Import spec files: import.spc

```{r UN_import2, echo = TRUE}
# import seasonal options from specification file

n3000013_import_spc <-  
    seasonal::import.spc("n3000013.spc")

names(n3000013_import_spc)

n3000013_import_spc$x

n3000013_import_spc$seas
```

## Import spec files: import.spc

```{r UN_import3, echo = TRUE}
# evaluate separate elements of n3000013_import_spc
library(seasonal)

eval(n3000013_import_spc$x)
n3000013_seas_from_file <- 
   eval(n3000013_import_spc$seas)

names(n3000013_seas_from_file)
```

## Running More Than One Series

- Can use list or mts (multiple time series) objects
- Run the `seas` function on the object
  - Old way: lists with the 'lapply' function
  - New way: process lists and mts objects directly

## Running More Than One Series

```{r UN_list, echo = TRUE}

UN_data_list <- 
   list(n3000013 = n3000013,
        n3000014 = seasonal::import.ts("data/n3000014.dat"),
        n3000025 = seasonal::import.ts("data/n3000025.dat"),
        n3000026 = seasonal::import.ts("data/n3000026.dat"))

#
# UN_data_list$n3000013 and
# UN_data_list[[1]] are equivalent
#
```

## Running More Than One Series

```{r UN_list2, echo = TRUE}

UN_airline_seas <- 
    seasonal::seas(UN_data_list, x11 = "", 
        check.print = c("none", "+acf", "+acfplot",
                        "+normalitytest"), 
        regression.aictest = NULL, 
        outlier.types = "all",
        arima.model = "(0 1 1)(0 1 1)")

names(UN_airline_seas)
```

## Running More than One Series

- Output is a list of seas objects
  - Elements have the same names as the list object
- To extract output or diagnostics, you can
  - Apply functions to the individual elements
  - Use `lapply` or `for` loops on the list as a whole

## Examples

- Get transformation information from list of objects
- Specify options for individual series using `list` argument
- Using mts objects with `seasonal`
- Composite seasonal adjustment with `seasonal`
- `out` argument
- User defined regressors with `seasonal`
- Saving regARIMA model summaries

## Transformation Choice

```{r UN_trans, echo = TRUE}
UN_names <- names(UN_data_list)

# get transformation choice, AICC 

UN_AICC_nolog  <- 
   lapply(UN_airline_seas, function(x)
          try(seasonal::udg(x, "aictest.trans.aicc.nolog")))
UN_AICC_log    <- 
   lapply(UN_airline_seas, function(x)
          try(seasonal::udg(x, "aictest.trans.aicc.log")))
UN_AICC_trans  <- 
   lapply(UN_airline_seas, function(x)
          try(seasonal::udg(x, "aictrans")))

# Generate difference between log and no log AICC
UN_AICC_diff   <- list()

for (i in 1:length(UN_names)) {
    this_series <- UN_names[i]
    UN_AICC_diff[[this_series]] <-
        UN_AICC_log[[this_series]] - UN_AICC_nolog[[this_series]]
}

```
## Transformation Choice

```{r UN_trans2, echo = TRUE}
# collect items into a data frame
UN_AICC_df <- 
    data.frame(nolog = unlist(UN_AICC_nolog), 
               log = unlist(UN_AICC_log), 
               diff  = unlist(UN_AICC_diff),  
               trans = unlist(UN_AICC_trans))
rownames(UN_AICC_df) <- UN_names

UN_AICC_df
```

## Using mts Objects

```{r UN_mts, echo = TRUE}

# An alternate approach - using mts objects

UN_data_mts <- 
  cbind(n3000013, 
        UN_data_list$n3000014, 
        UN_data_list$n3000025, 
        UN_data_list$n3000026)
colnames(UN_data_mts) <- names(UN_data_list)

UN_airline_mts_seas <- 
    seasonal::seas(UN_data_mts, x11 = "", 
        check.print = c("none", "+acf", "+acfplot",
                        "+normalitytest"), 
        regression.aictest = NULL, 
        outlier.types = "all",
        arima.model = "(0 1 1)(0 1 1)")

names(UN_airline_mts_seas)
```

## Specifying options for each series

```{r UN_individual, echo = TRUE}
UN_individual_seas <- 
  seasonal::seas(UN_data_mts,
    x11 = "", 
    check.print = c("none", "+acf", "+acfplot", "+normalitytest"), 
    regression.aictest = NULL, 
    outlier.types = "all",
    arima.model = "(0 1 1)(0 1 1)",
    list = list(
        list(transform.function = "none"),
        list(transform.function = "log"),
        list(transform.function = "log"),
        list(transform.function = "log")
    )
)
names(UN_individual_seas)

```

## Another alternate approach

```{r UN_individual2, echo = TRUE}
# An alternate approach - specifiying time series within the list

UN_auto_x11_seas <- 
  seasonal::seas(
    check.print = c("none", "+acf", "+acfplot", "+normalitytest"), 
    regression.aictest = NULL, 
    outlier.types = "all",
    x11.seasonalma = "s3x5",
    list = list(
        list(x = UN_data_list$n3000013, transform.function = "none"),
        list(x = UN_data_list$n3000014, transform.function = "log"),
        list(x = UN_data_list$n3000025, transform.function = "log"),
        list(x = UN_data_list$n3000026, transform.function = "log")
    )
)

names(UN_auto_x11_seas)

names(UN_auto_x11_seas) <- names(UN_data_list)
names(UN_auto_x11_seas)
```

## Composite seasonal adjustment

- Two components - A and B
- Indirect adjustment: SA(A) + SA(B)
- Direct adjustment: SA(A + B)

## Composite seasonal adjustment in seasonal package

```{r UN_composite, echo = TRUE}
UN_composite_seas <- seasonal::seas(UN_data_mts,
     x11 = "", 
     check.print = c("none", "+acf", "+acfplot", "+normalitytest"), 
     regression.aictest = NULL, 
     outlier.types = "all",
     arima.model = "(0 1 1)(0 1 1)",
     series.comptype = "add",
     list = list(
         list(transform.function = "none"),
         list(transform.function = "log"),
         list(transform.function = "log"),
         list(transform.function = "log")
     ),
     composite = 
       list(arima.model = "(0 1 1)(0 1 1)", x11 = "",
            check.print = c("none", "+acf", "+acfplot", "+normalitytest"),
       regression.aictest = NULL, 
       outlier.types = "all")
)

names(UN_composite_seas)
```

## A Note About Composite Adjustment

- Some `seasonal` functions do not have access to `composite` element
  - out
  - series
- Will be in the next build of seasonal
  
## The out argument

```{r UN_out1, echo = TRUE}
UN_composite_out_seas <- seasonal::seas(UN_data_mts,
     x11 = "", 
     check.print = c("none", "+acf", "+acfplot", "+normalitytest"), 
     regression.aictest = NULL, 
     outlier.types = "all",
     arima.model = "(0 1 1)(0 1 1)",
     series.comptype = "add",
     list = list(
         list(transform.function = "none"),
         list(transform.function = "log"),
         list(transform.function = "log"),
         list(transform.function = "log")
     ),
     composite = 
       list(arima.model = "(0 1 1)(0 1 1)", x11 = "",
            check.print = c("none", "+acf", "+acfplot", "+normalitytest"),
       regression.aictest = NULL, 
       outlier.types = "all"),
     out = TRUE
)

tf <- tempfile(fileext = ".html")
writeLines(UN_composite_out_seas$composite$out, tf)
browseURL(tf)
```

## The out argument

```{r UN_out2, echo = TRUE}
m_air_out <- seasonal::seas(AirPassengers,
                    x11="", out = TRUE,
                    arima.model="(0 1 1)(0 1 1)")

names(m_air)
names(m_air_out)

object.size(m_air)
object.size(m_air_out)

head(m_air_out$out, n=10)
```
## Adding User Defined Regressors

```{r UN_userreg, echo = TRUE}
start_reg <- c(2003, 1)
end_reg   <- c(2030,12)
tc_2020_apr_p9 <- 
    sautilities::gen_tc_outlier_ts(c(2020,4), start_reg, end_reg, 
                                   tc_alpha = 0.9)
tc_2020_apr_p7 <- 
    sautilities::gen_tc_outlier_ts(c(2020,4), start_reg, end_reg, 
                                   tc_alpha = 0.7)

n3000013_final_seas_tc_p9 <- 
    seasonal::seas(x = n3000013, xreg = tc_2020_apr_p9, 
                   regression.usertype = c("ls"), 
                   transform.function = "none", 
                   regression.variables = c("ls2008.may", "ls2021.Apr"), 
                   arima.model = "(0 1 1)(0 1 1)", 
                   estimate.exact = "arma", estimate.maxiter = 10000, 
                   forecast.maxlead = 60, 
                   check.print = c("none", "+acf", "+acfplot", 
                                   "+normalitytest"), 
                   seats.save = c("s14", "mdc"), 
                   regression.aictest = NULL, 
                   outlier = NULL)

n3000013_final_seas_tc_p7 <- 
    seasonal::seas(x = n3000013, xreg = tc_2020_apr_p7, 
                   regression.usertype = c("ls"), 
                   transform.function = "none", 
                   regression.variables = c("ls2008.may", "ls2021.Apr"), 
                   arima.model = "(0 1 1)(0 1 1)", 
                   estimate.exact = "arma", estimate.maxiter = 10000, 
                   forecast.maxlead = 60, 
                   check.print = c("none", "+acf", "+acfplot", 
                                   "+normalitytest"), 
                   seats.save = c("s14", "mdc"), 
                   regression.aictest = NULL, 
                   outlier = NULL)

n3000013_tc_AICC <- 
    as.numeric(c(seasonal::udg(n3000013_final_seas_tc_p9, "aicc"), 
                 seasonal::udg(n3000013_final_seas_tc_p7, "aicc")))
n3000013_tc_AICC
```

## Plot of ori, SA, trend

```{r UN_plot2, echo = TRUE, dev='png'}
this_x            <- window(n3000013_final_seas_tc_p7$x, start = c(2016, 1))
n3000013_sa_p7    <- seasonal::final(n3000013_final_seas_tc_p7)
this_sa_p7        <- window(n3000013_sa_p7, start = c(2016, 1))
n3000013_trend_p7 <- seasonal::trend(n3000013_final_seas_tc_p7)
this_trend_p7     <- window(n3000013_trend_p7, start = c(2016, 1))
plot(this_x, main = "n3000013: SA and Trend", ylab = "n3000013",
     col = "grey")
lines(this_sa_p7, col = "blue")
lines(this_trend_p7, col = "brown")
```

## Save regARIMA model summaries

```{r UN_regarima, echo = TRUE, dev='png'}
sautilities::get_regarima_estimates_matrix(n3000013_final_seas_tc_p7,
                                           add_diff = TRUE,
                                           this_xreg_names = "TC2020.4 (alpha=0.7)")

sautilities::get_regarima_estimates_matrix(n3000013_final_seas_tc_p9,
                                           add_diff = TRUE,
                                           this_xreg_names = "TC2020.4 (alpha=0.9)")
```

## References

- Seasonal Adjustment by X-13ARIMA-SEATS in R by Chistophe Sax and Dick Eddelbuettel
  - Excellent introduction
  - Details of a graphical user interface
  - Large scale production use
  - `x13binary` package
  
<https://cran.r-project.org/web/packages/seasonal/vignettes/seas.pdf>
  
## References

- Adjusting Multiple Series
  - Covers all aspects of the topic
  - Details of the multimode argument
  
<https://cran.r-project.org/web/packages/seasonal/vignettes/multiple.html>

## References

- Examples of X-13ARIMA SEATS in R
  - Gives every example from the X-13ARIMA-SEATS Manual
- For each example
  - X-13 code
  - Equivalent seasonal code
  - Remarks on the implementation
  - Allow you to run it yourself

<http://www.seasonal.website/examples.html>

## Thank you for your time

Brian C. Monsell\
OEUS\
monsell.brian@bls.gov

## Appendix: seas Object Structures

- Input structures
- Data structures
- Model structures


## Input, seas object

`n3000013_seas$spc`

- Spec file used in the seas run
- Spec input is stored as a list

`n3000013_seas$call`

- Preserves the call to seasonal

## input structures, seas object

```{r UN_spc_structure, echo = TRUE}
names(n3000013_seas$spc)

n3000013_seas$spc$regression

n3000013_seas$spc$arima

```

## input structures, seas object

```{r UN_call_structure, echo = TRUE}
n3000013_seas$call
```

## Data structures, seas object

```{r UN_data1_structure, echo = TRUE}
names(n3000013_seas$series)

head(n3000013_seas$series$s10)
```

## Data structures, seas object

```{r UN_data2_structure, echo = TRUE}
colnames(n3000013_seas$data)

head(n3000013_seas$data[,2])
```
## Data structures, seas object

```{r UN_data3_structure, echo = TRUE}
tail(n3000013_seas$x)
```
## Seas Object: model

```{r UN_model, echo = TRUE}
names(n3000013_seas$model)
n3000013_seas$model$regression
n3000013_seas$model$arima
```

## Seas Object: est

```{r UN_est1, echo = TRUE}
names(n3000013_seas$est)

n3000013_seas$est$reg
n3000013_seas$est$reg$estimate
as.numeric(n3000013_seas$est$reg$estimate)
```
